---
title: "Data Wrangling"
format: html
---

Code chunk shortcut:
Mac: Option + Command + I
Windows: Ctrl + Alt + I 
```{r}
#| message: false
library(dplyr)
library(tidyr)
library(readr)
```

## Data Sources

### Read in the data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```
### Explore data

```{r}
# prints the column names 
colnames(catch_original)

# first 6 lines of the data
head(catch_original)

# summary of each column
summary(catch_original)

# unique values in column
unique(catch_original$Region)

# open data frame
#View(catch_original)
```

## Clean and Reshape

Shortcut for %>%:
Windows: Ctrl + Shift + M
Mac: cmd + shift + M

### Using select() function
```{r}
catch_data <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)
```

```{r}
catch_data <- catch_original %>%
  # drop the All and notesRegCode columns
  select(-All, -notesRegCode)
```

### Quality check
```{r}
dplyr::glimpse(catch_data)
```

### Changing column content using mutate()
```{r}
catch_clean <- catch_data %>%
  # change column type to numeric
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)
```

```{r}
i <- which(is.na(catch_clean$Chinook))
i
```

```{r}
catch_data[i,]
```
```{r}
catch_clean <- catch_data %>%
  # if Chinook column is I, then change it to 1
  # otherwise leave it 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         # also change the Chinook column type to numeric
         Chinook = as.numeric(Chinook))

catch_clean[i,]
```
### Changing shape
```{r}
catch_long <- catch_clean %>%
  # pivot longer all columns except Region and Year
  pivot_longer(
    cols = -c(Region, Year),
    names_to = "species",
    values_to = "catch"
  )

head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
  # pivot wider to make more salmon columns
  pivot_wider(names_from = species,
              values_from = catch)

head(catch_wide)
```

### Renaming columns
```{r}
catch_long <- catch_long %>%
  rename(catch_thousands = catch)

head(catch_long)
```

### Adding columns
```{r}
catch_long <- catch_long %>%
  # create a new catch column that has fish in the correct units
  mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

head(catch_long)
```

### Summary statistics
```{r}
unique(catch_original$Region)
```

```{r}
mean_region <- catch_long %>%
  # group by each region
  group_by(Region) %>%
  # calculate mean catch for each region
  summarize(mean_catch = mean(catch))

head(mean_region)
```

### Filtering rows using filter()
```{r}
sse_catch <- catch_long %>%
  filter(Region == "SSE")

head(sse_catch)
```

```{r}
catch_millions <- catch_long %>%
  # filter to catches of over 1 million fish
  filter(catch > 1000000)

head(catch_millions)

chinook_sse <- catch_long %>%
  # filter to just Chinook from SSE
  filter(Region == "SSE" & species == "Chinook")

sse_nse <- catch_long %>%
  # filter to SSE or NSE
  filter(Region == "SSE" | Region == "NSE")
```
### Sort your data
```{r}
mean_region <- catch_long %>%
  # group by each region
  group_by(Region) %>%
  # calculate mean catch for each region
  summarize(mean_catch = mean(catch)) %>%
  # sort by mean catch
  arrange(desc(mean_catch))

head(mean_region)
```
### Split and unite columns
```{r}
# example data
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))

sites_df %>%
  separate(col = site, 
           into = c("island", "site_number"),
           sep = "-")
```
```{r}
## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))

cities_df %>%
  separate(city, c("city", "state_code"), " ")
```

```{r}
# example data
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)

dates_df %>%
  unite(date, year, month, day, sep = "-")
```
## Now all together!
```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

